---
title: "Class Reference"
author: "Michael Howes"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Consider this a personal guide to the commands and functions you will learn. In general, when you come across an R command or function that you want to remember, put it in here along with a description of what it does and when you'd use it.

### How to set the working directory

The command to set a working directory is setwd(). For example, this sets the working directory to a directory called data_journalism_2024_fall inside a directory called jour472 located inside your home directory (the ~ is equivalent of /Users/[youruser] on a Mac).

```{r}
setwd("~/jour472/data_journalism_2024_fall")
```

### How to install and load the tidyverse

```{r}
install.packages("tidyverse")
library(tidyverse)
```

library(readr)
library(dplyr)

These two commands import the readr and dplyr libraries. THe library readr reads different types of data while dplyr analyzes the data. 

How to import data --- EXAMPLE

Marland_winred_contributions <-read_rds("maryland_winred.rds")

Variable: maryland_winred_contributions 
  - Variables are the name to refer to complex things, in this case being the data imported into R to serve as a dataframe. 

Variable assignment operator: <-
  - Assigns something to a word
  
Arguments: What is inside the read_rds() function
- the name of the file that wats to be loaded


sum(): Takes the sum of all the values

summarize(): Inside the parentheses in summarize,  set up the summaries that are wanted. 

arrange (): Arranges data in order. In ascending order by default, but can by sorted in descending order.