---
title: "lab_03"
author: "Michael Howes"
date: "2024-09-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse
library(tidyverse)

# Load janitor
library(janitor)
```

## Load data

```{r}
# Load MD slice of FEC House candidate contribution data
md_contribs <- read_rds("data/lab_03.rds")
```

## Fix header issues

You'll need to fix one issue with the column names using janitor. Write a code block to do that, making sure that you save the results to the same dataframe.

```{r}
clean_md_contribs <- md_contribs |>
  clean_names() |>
  mutate(contributor_zip = str_sub(contributor_zip, start = 1L, end = 5L)) |>
  mutate(contributor_city = str_to_title(contributor_city)) |>
  mutate(
    city = case_when(
      contributor_city %in% c("Balrimore", "Blatimore") ~ "Baltimore",
      contributor_city %in% c("Anapolis") ~ "Annapolis",
      contributor_city %in% c("Potamic") ~ "Potomac",
      contributor_city %in% c("4615 North Park Ave Apt. 1410") ~ "Chevy Chase",
       contributor_city %in% c("Silver Springs") ~ "Silver Spring",
      TRUE ~ contributor_city
    )
  )
```


## Answer questions

Q1: Which seven cities in Maryland had the largest *number* of contributions from donors? How many did each have? To answer this question, you'll have to clean the messy "contributor_city" field.

Note: To answer this question, you'll need to see every variation of the "contributor_city" for the top seven cities, which you should identify through some initial analysis. Then find variations of them and change those variations into the most common version.

A1:Bethesda had the most contributions with 1,177. Chevy Chase was next with 705 contributions. Baltimore and Annapolis were next, with 692 and 491 respectively. Potomac was No. 5 with 491 contributions and Silver Spring had 408 contributions. Lastly, Rockville was No. 7 with 358 contributions. 


```{r}
top_cities <- clean_md_contribs |>
  group_by(city) |>
  summarise(
    count = n()
  ) |>
  arrange(desc(count))
```


Q2: Find the 5-digit ZIP code in Maryland that had the highest total amount of contributions (by dollar value). You will need to use mutate to ensure that you have a zip code column with the first five characters from the existing zip code column.

Where is that zip code and what kind of area is it, demographically? Which committee is the largest recipient of money from that zip code? How many of the top five recipients are Maryland candidates?

A2: The zip code 20817 had the highest ammount of total contributions with $731,489.80. That zip code is located in Bethesda, Md. This makes sense as Bethesda is ranked as the No. 1 city in Maryland in terms of median household income with $166,380, according to homearea.con. Helmer for Congress is the largest recipient of money from the zip code, a previous democratic candidate for the House from Virginia. Only two committees in the top five recipients are Maryland candidates, Jamie Raskin and April McClain Delaney. 

```{r}
zip_contributions <- clean_md_contribs |>
  group_by(contributor_zip) |>
  summarise(total_contribution = sum(contribution_receipt_amount)) |>
  arrange(desc(total_contribution))
```

```{r}
zip_contributions_committee <- clean_md_contribs |>
  filter(contributor_zip == "20817") |>
  group_by(contributor_zip, committee_name) |>
  summarise(total_contribution = sum(contribution_receipt_amount)) |>
  arrange(desc(total_contribution))
```

Q3: Let's look at top individual donors. Write code to generate a list of top donors using name and city, adding up the total amount of contributions. What does the original data say about the top two donors - who are they and what do they do? Do you see anything in the list of top 10 donors that would make you uncertain about the accuracy of these results?

A3:The top contributors in Maryland both have the last name Jacobsen and are from Bethesda. I then checked and saw that they also live in the both address --- leading me to believe they are a married couple. Mark Jacobsen works in banking services while Melanie Jacobsen is listed as a farmer. One thing that would make me uncertain is that there are a couple duplicated listed, the Jacobsens, Pughs, and Drezners. 

```{r}
top_donors <- clean_md_contribs |>
  group_by(contributor_name, contributor_city, contributor_street_1) |>
  summarise(total_contribution = sum(contribution_receipt_amount)) |>
  arrange(desc(total_contribution))
```
