---
title: "lab_01"
author: "Michael Howes"
date: "2024-09-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Turn off scientific notation
options(scipen=999)
```

## About this lab

To complete this lab, you need to: \* run existing code as directed (look for **Task**). \* modify existing code as directed (look for **Task**). \* write code in empty codeblocks provided to answer questions included (look for **Q**). \* write out the answer in the form of a complete sentence in the space given (look for **A**). When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

Remember to add any new verbs or concepts to your reference notebook!

## Load libraries and establish settings

**Task** Run the codeblock below to load the Tidyverse. To run the codeblock, hit the little green play button in the upper right corner of the gray area (the codeblock) below.

```{r}
# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
```

## Load Data

We're going to load the slice of WinRed contribution data from West Virginia, which is in a file called `lab_01.rds`

**Task** Using the example from the pre_lab, write code to load the RDS file and assign it to a good variable name.

```{r}
westvirginia_winred_contributions <- read_rds("lab_01.rds")
```

## Examine Data

**Task** Run the codeblock below to use glimpse to get a sense of the column names, types and sample data.

**Q1** How many rows and columns does this dataframe have?

**It has 40,441 rows and 24 columns.**

```{r}
glimpse(westvirginia_winred_contributions)
```

## Answer questions

**Q2** Which West Virginia city provided the largest number of contributions? How many?

**The city of Charleston provided the largest number of contributions in West Virginia with 3,226 contributions.**

```{r}
westvirginia_winred_contributions |>
  group_by(city) |>
  summarise(
    count_contribs = n(),
    total_amount = sum(amount),
    mean_amount = mean(amount),
    median_amount = median(amount),
    min_amount = min(amount),
    max_amount = max(amount)
  ) |>
   arrange(desc(count_contribs))
```

**Q3** Which West Virginia cities accounted for the fewest contributions? How many of them, roughly speaking?

**There were 63 different cities in West Virginia which were shown to have just one contribution.**

```{r}
westvirginia_winred_contributions |>
  group_by(city) |>
  summarise(
    count_contribs = n(),
    total_amount = sum(amount),
    mean_amount = mean(amount),
    median_amount = median(amount),
    min_amount = min(amount),
    max_amount = max(amount)
  ) |>
   arrange(desc(count_contribs))
```

**Q4** What's your best educated guess to explain why the city identified in question 2 had more contributions compared to the cities identified in question 3? Hint: you won't find the answer in the data, you'll need to do some quick web research. You do not need to write any code for this question, so no codeblock is provided.

**Charleston has the largest population of any city in West Virginia with 46,237 residents in 2024. The cities with only one contribution have less residents, such as the city of Milton which has 2,738 residents according to worldpopulationreview.com.  **

**Q5** What recipient got the largest total amount of money? How many contributions did they receive and what was the average donation?

**The Trump National Committee JFC recieved the largest amount of money with $463,944. It recieved 14,063 total contributions with an average amount of $32.99.**

```{r}
westvirginia_winred_contributions |>
  group_by(memo_text) |>
  summarise(
    count_contribs = n(),
    total_amount = sum(amount),
    mean_amount = mean(amount),
    median_amount = median(amount)
  ) |>
  arrange(desc(total_amount))
```

**Q6** What is that top recipient and what explains its popularity? Hint: this will require web research, you won't be able to answer this purely from the data set. You do not need to write any code for this question, so no codeblock is provided.

**The Trump National Committee JFC is a joint fundraising committee between Donald Trump and the Republican National Committee. It is more popular than the other recipients because of that combination, as individuals who are both Trump supporters and registered Republicans, a large group, likely donate to it. **

**Q7**. Using the answer from A5 as a starting point, produce a summary table that has the following information for each recipient: \* count of contributions \* total amount of all contributions \* the average contribution amount \* the median contribution amount \* the smallest contribution amount \* the largest contribution amount

Store it as a variable object called `wv_summary` so that you can examine it with the data viewer. What newsworthy nuggets jump out at you that you could imagine dropping into a news story? Write a few summary sentences with overall findings.

**I found it interesting that Carol D. Miller was receiving an average of $1,490 on 15 contributions. Her median contribution was $1,000, tied for third most. I also found that the second-most total contributions went to refunds, which is interesting. Lastly, I caught that seven different groups recieved totals that were less than a dollar, including the MAJORITY COMMITTEE PAC which recieved six donations. I noted that all committees for congresspeople in that group were not from the state of West Virginia.    **.

```{r}
wv_summary <- westvirginia_winred_contributions |>
  group_by(memo_text) |>
  summarise(
    count_contribs = n(),
    total_amount = sum(amount),
    mean_amount = mean(amount),
    median_amount = median(amount)
  ) |>
  arrange(desc(total_amount))
```
